---
title: "Inclass_Assignment_MLR"
author: "S20426"
date: "2025-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### 1 Load the Dataset

```{r}
library(AppliedPredictiveModeling)
data(abalone)
attach(abalone)
```

### 2 Graphical Interpretation

#### Draw a correlogram using corrplot() function

```{r}
# install.packages("corrplot")
library(corrplot)
corrplot(cor(abalone[,-1]), tl.col = "red")
```



####  Draw scatter plots using chart.Correlation():

```{r include=FALSE}
# install.packages("PerformanceAnalytics")
library("PerformanceAnalytics")
```

```{r}
# install.packages("PerformanceAnalytics")
# library("PerformanceAnalytics")
chart.Correlation(abalone[,-1], histogram=TRUE, pch=19)
```


###  02. Parameter Estimation

#### Check For Missing Values

```{r}
sum(is.na(abalone))

```


```{r}
model1 <- lm(Rings ~ as.factor(Type) + LongestShell + Diameter + 
             WholeWeight + ShuckedWeight + VisceraWeight + ShellWeight, 
             data = abalone)
model1
```

#### Check Model Summary

```{r}
summary(model1)

```



#### Create 2 columns For Type

```{r}
abalone$Type.I=ifelse(abalone$Type=="I",1,0)
abalone$Type.M=ifelse(abalone$Type=="M",1,0)
head(abalone)
```

```{r}
full_model <-  lm(Rings~Type.I + Type.M+ LongestShell+ Diameter+ WholeWeight+ ShuckedWeight+ VisceraWeight+ ShellWeight,data=abalone)
summary(full_model)
```


### 03 Test on Parameters

```{r}
library(MASS)
step_model <- stepAIC(full_model, direction = "both")
summary(step_model)

```

### 04 Analysis of Variance


#### check model assumption

```{r}
par(mfrow =c(2,2))
plot(step_model)
```

#### Identifying Influential Outliers

```{r}
standard_res=rstandard(step_model)
plot(fitted(step_model), standard_res)
```


#### Studentized Residuals

```{r}
student_res=rstudent(step_model)
plot(fitted(step_model), student_res)
```

#### Variance inflation factor

```{r}
library(car)
vif_values <- vif(step_model)          
barplot(vif_values, main = "VIF Values", horiz = TRUE, col = "steelblue") 
abline(v = 5, lwd = 3, lty = 2)    

```

### 05 Interpretation and Prediction

#### Testing the multicollinearity among the predictor variables

```{r}
library(mctest)
omcdiag(mod=step_model)     

# OR

#imcdiag(mod=step.model)   

```
#### model summary()

```{r}
summary(step_model)
```


