---
title: "Multiple_Linear_Regression"
author: "S20426"
date: "2025-03-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## How to fit a multiple linear regression model?


### Steps you can follow:

#### 01.Graphical Interpretation
#### 02.Parameter Estimation
#### 03.Tests on Parameters
#### 04.Analysis of Variance
#### 05.Interpretation and Prediction


##                    01.Graphical Interpretation

```{r}

```


```{r}
# install.packages("AppliedPredictiveModeling")
library(AppliedPredictiveModeling)
data(abalone)
attach(abalone)
# head(abalone)
# summary(abalone)
# cor(abalone[,-1]) correlation matrix
```

###  Draw a correlogram using corrplot() function

```{r}
# install.packages("corrplot")
library(corrplot)
corrplot(cor(abalone[,-1]), tl.col = "red")
```

###  Draw scatter plots using chart.Correlation():

```{r include=FALSE}
# install.packages("PerformanceAnalytics")
library("PerformanceAnalytics")
```


```{r}
# install.packages("PerformanceAnalytics")
# library("PerformanceAnalytics")
chart.Correlation(abalone[,-1], histogram=TRUE, pch=19)
```

##                    02. Parameter Estimation

#### Check For Missing Values

```{r}
sum(is.na(abalone))

```


```{r}
model1 <- lm(Rings ~ as.factor(Type) + LongestShell + Diameter + 
             WholeWeight + ShuckedWeight + VisceraWeight + ShellWeight, 
             data = abalone)
model1
```

#### Check Model Summary

```{r}
summary(model1)

```
```{r}
head(abalone)
```
#### Create 2 columns For Type

```{r}
abalone$Type.I=ifelse(abalone$Type=="I",1,0)
abalone$Type.M=ifelse(abalone$Type=="M",1,0)
head(abalone)
```

```{r}
model2 <-  lm(Rings~Type.I + Type.M+ LongestShell+ Diameter+ WholeWeight+ ShuckedWeight+ VisceraWeight+ ShellWeight,data=abalone)
summary(model2)
```
#  Methods of building a multiple linear regression model

### 01 Backward Elimination
### 02 Forward Selection
### 03 Bidirectional Elimination (Stepwise regression)

## 01 Backward Elimination


#### 1. Set a significance level for which data will stay in the model.

#### 2. Next, fit the full model with all possible predictors.

#### 3. Consider the predictor with the highest P-value. If the P-value is greater than the significance level, you’ll move to step four, otherwise, you’re done!

#### 4. Remove that predictor with the highest P-value.

#### 5. Fit the model without that predictor variable.

#### 6. Go back to step 3, do it all over, and keep doing that until you come to a point where even the highest P-value is < SL.

### I.Step() Function (AIC-Based Backward Elimination)


* Uses Akaike Information Criterion (AIC) to decide which variables to remove.
* AIC balances model fit and complexity:
* Lower AIC = Better trade-off between accuracy and simplicity.
* Stepwise removal continues until removing a variable increases AIC.
* Can retain variables with high p-values if they improve overall model performance.


```{r}
model2 <-  lm(Rings~Type.I + Type.M+ LongestShell+ Diameter+ WholeWeight+ ShuckedWeight+ VisceraWeight+ ShellWeight,data=abalone)
step.model <- step(model2, direction = "backward",trace=FALSE)
summary(step.model)
```

### II.drop1SignifReg() (Significance-Based Backward Elimination)

* Drops one variable at a time based on p-value until only significant variables (p < 0.05)   remain.
* Uses hypothesis testing to justify variable inclusion.

* More statistically rigorous but may remove useful variables if their effect is small.

```{r}
library(SignifReg)
nullmodel = lm(Rings ~ 1, data=abalone)  
fullmodel = lm(Rings ~ Type.I + Type.M + LongestShell + Diameter + WholeWeight + ShuckedWeight + VisceraWeight + ShellWeight, data=abalone)

```


```{r}
scope = list(lower=formula(nullmodel), upper=formula(fullmodel))
Sub.model = drop1SignifReg(fullmodel, scope, alpha = 0.05, criterion = "p-value", override = "TRUE")

# This function iteratively removes the least significant predictor (highest p-value) until all remaining predictors are significant at alpha = 0.05.
# criterion = "p-value" ensures the variable removal is based on significance.
# override = "TRUE" forces the function to proceed even if some assumptions are not fully met.
summary(Sub.model)


```

* Type.M (p = 0.597) and LongestShell (p = 0.977) have very high p-values.
* Normally, these should be removed.
* However, they might remain due to interactions, multicollinearity, or their effect on the overall model stability.
* The function ensures that the model performs well as a whole rather than strictly eliminating variables with high p-values.


## 02 Forward Selection

1. Choose the significance level (SL = 0.05).
2. Fit all possible simple regression models and select the one with the lowest P-value.
3. Keep this variable and fit all possible models with one extra predictor added to the one you already have.
4. Find the predictor with the lowest P-value. If P < Sl, go back to step 3. Otherwise, you’re done!

### I Forward selection using SignifReg package

```{r}
library(SignifReg)
nullmodel = lm(Rings~1, data=abalone)
fullmodel = lm(Rings~Type.I + Type.M+ LongestShell+ Diameter+ WholeWeight+ ShuckedWeight+ VisceraWeight+ ShellWeight, data=abalone)
scope = list(lower=formula(nullmodel),upper=formula(fullmodel))

sub.model2=add1SignifReg(nullmodel,scope, alpha = 0.05, criterion = "p-value",override="TRUE")
summary(sub.model2)
```
### II Stepwise Regression


1.Select a significance level to enter and a significance level to stay.

2.Perform the next step of forward selection where you add the new variable.

3.Now perform all of the steps of backward elimination.

4.Now head back to step two, then move forward to step 3, and so on until no new variables can enter and no new variables can exit.
```{r}
model2 <-  lm(Rings~Type.I + Type.M+ LongestShell+ Diameter+ WholeWeight+ ShuckedWeight+ VisceraWeight+ ShellWeight,data=abalone)
step.model2 <- step(model2, direction = "forward", trace=FALSE)
summary(step.model2)
```
###  Backward Elimination (Stepwise Backward Selection)
 Starts with: The full model (all predictors included).
 Process:

Removes the least significant predictor (highest p-value) one at a time.
Reassesses the model after each removal.
Stops when all remaining predictors are statistically significant (p-value < α, usually 0.05).

#### Advantages:
nsures that we do not miss important predictors.
tarts with the best-fitting model (high R²) and simplifies it.

#### Disadvantages:
Can keep irrelevant predictors if they interact with others.
omputationally expensive for large datasets.

###  Forward Selection (Stepwise Forward Selection)
Starts with: The null model (only intercept, no predictors).
Process:

Adds the most significant predictor (lowest p-value or highest contribution to model improvement).
Continues adding variables one at a time based on statistical significance.
Stops when adding another variable does not significantly improve the model (p-value > α).

#### Advantages:
tarts with a simple model and gradually builds complexity.
ess risk of overfitting.

#### Disadvantages:
ght miss important predictors that are only significant when combined with others.
lower than backward elimination because it starts from scratch.


#   Model Diagnosis


