---
title: "Hypothesis_Testing"
author: "S20426"
date: "2025-03-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##                     01 Propertion Testing
### One Sampling

#### Large Sample

```{r}
set.seed(10)
 Hair_col <- c(rep("black", 1500), rep("brown", 1000), rep("blonde", 500))
 sampleP <- sample(Hair_col,1000)
 Ptable <-table(sampleP)         
 prop.test(x = 498, n= 1000, p=0.5, alternative = "two.sided", conf.level = 0.95, correct = FALSE)

```

#### Small Sample

```{r}
set.seed(10)
sampleS<- sample(Hair_col,10)
Stable <- table(sampleS)
Stable

binom.test(x= 3, n=10, p=0.5, alternative = "two.sided" )   
```

### Two Sampling

```{r}
prop.test(x=c(490,400),n=c(500,500))
? prop.test()

```

##                     02 Mean Testing

### One Sample

#### One Sample Test for Mean: when population variance is known.

```{r}
weights=c(165.1,171.5,168.1,165.6,166.8,170.0,168.8,171.1,168.8,173.6,163.5,169.9,165.4,174.4,171.8,166.0,174.6,174.5,166.4,173.8)

library("BSDA")
z.test(x=weights, mu = 160,sigma.x =15, alternative = "two.sided")
```

#### One Sample Test for Mean: when population variance is unknown.

```{r}
weights <- c(165.1,171.5,168.1,165.6,166.8,170.0,168.8,171.1,168.8,173.6,163.5,169.9,165.4,174.4,171.8,166.0,174.6,174.5,166.4,173.8)
t.test(weights, mu = 160, alternative = "two.sided")

```

###  Two Sample 

#### Case 1 : Equal Variances

```{r}
fat_m <- c(13.3,6.0,20.0,8.0,14.0,19.0,18.0,25.0,16.0,24.0,15.0,1.0,15.0)
fat_w <- c(22.0,16.0,21.7,21.0,30.0,26.0,12.0,23.2,28.0,23.0)
t.test(fat_w,fat_m, var.equal = TRUE)

```
#### Case 2 : Unequal Variances

```{r}
fat_m <- c(13.3,6.0,20.0,8.0,14.0,19.0,18.0,25.0,16.0,24.0,15.0,1.0,15.0)
fat_w <- c(22.0,16.0,21.7,21.0,30.0,26.0,12.0,23.2,28.0,23.0)

t.test(fat_w,fat_m, var.equal = FALSE)
```
#### Paired Sample

The mean of paired samples refers to the mean difference between two related (dependent) samples. This is typically used in paired t-tests or before-and-after studies, where each observation in one sample has a corresponding observation in the other sample.

```{r}
lab1 <- c(22,18,28,26,13,8,21,26,27,29,25,24,22,28,15)
lab2 <- c(25,21,31,27,11,10,25,26,29,28,26,23,22,25,17)

t.test(lab1,lab2, paired = TRUE)
```


##                     03 Testing For Variance

### One sample

```{r}
weights <- c(165.1,171.5,168.1,165.6,166.8,170.0,168.8,
             171.1,168.8,173.6,163.5,169.9,165.4,174.4,
             171.8,166.0,174.6,174.5,166.4,173.8)

library(EnvStats)
varTest(weights, sigma =10, alternative = "two.sided")
```

### Two sample

```{r}
fat_m <- c( 13.3,6.0,20.0,8.0,14.0,19.0,18.0,25.0,16.0,24.0,15.0,1.0,15.0)
fat_w <- c(22.0,16.0,21.7,21.0,30.0,26.0,12.0,23.2,28.0,23.0)

var.test(fat_m, fat_w)
```

##                     04 Check the Normality Assumption

### Shapiro-Wilk test

```{r}
# shapiro.test(my_data)
```


### Anderson-Darling test

```{r}
# install.packages('nortest')
# library(nortest)
# ad.test(my_data)
```

### Kolmogorov-Smirnov test

```{r}
# ks.test(my_data, "pnorm")
```




##                  05. Sign Test

The Sign Test is a non-parametric statistical test used to determine whether there is a significant difference between the median of a sample and a hypothesized median, or to compare two paired samples. Unlike parametric tests, it does not assume normality and is useful when dealing with ordinal or skewed data.


#### When to Use the Sign Test?
One-Sample Sign Test: Used to test if the median of a single sample is equal to a hypothesized value.
Paired-Sample (Two-Sample) Sign Test: Used for paired data (dependent samples) to determine if there is a difference in the medians.


### Case 1 : Sign Test For  One Sample

```{r}
m0 <- 130 # median in 2008
prices <- c(230.00,148.00,126.00,134.62,155.00,157.70,
            160.00,225.00,125.00,109.00,157.00,115.00,
            125.00,225.00,118.00,179.00,176.00,125.00,
            123.00,180.00,151.00,120.00,143.00,170.00,
            190.00,233.00,148.72,189.00,121.00,149.00,
            225.00,240.00)

library(BSDA)
SIGN.test(prices,md = m0, alternative="greater")
```

### Case 1 :Sign Test for One Sample


```{r}
lab1 <- c(22,18,28,26,13,8,21,26,27,29,25,24,22,28,15)
lab2 <- c(25,21,31,27,11,10,25,26,29,28,26,23,22,25,17)

library(BSDA)
SIGN.test(x = lab1,
          y = lab2,
          alternative = "two.sided",
          conf.level = 0.95)
```

##                  06. Wilcoxon Sign Rank Test

### Case I: One Sample Sign Rank Test

```{r}
m0 <- 130 # median in 2008
prices <- c(230.00,148.00,126.00,134.62,155.00,157.70,
            160.00,225.00,125.00,109.00,157.00,115.00,
            125.00,225.00,118.00,179.00,176.00,125.00,
            123.00,180.00,151.00,120.00,143.00,170.00,
            190.00,233.00,148.72,189.00,121.00,149.00,
            225.00,240.00)

wilcox.test(prices,m0, exact = FALSE, alternative = "greater")
```

### Case II: Sign Rank Test for Two Paired Samples

```{r}
lab1 <- c(22,18,28,26,13,8,21,26,27,29,25,24,22,28,15)
lab2 <- c(25,21,31,27,11,10,25,26,29,28,26,23,22,25,17)

wilcox.test(lab1, lab2, paired = TRUE, exact = FALSE)

```
### Case III: Sign Rank Test for Two independent Samples

```{r}
fat_m <- c( 13.3,6.0,20.0,8.0,14.0,19.0,18.0,25.0,16.0,24.0)
fat_w <- c(22.0,16.0,21.7,21.0,30.0,26.0,12.0,23.2,28.0,23.0)
wilcox.test(fat_m, fat_w, alternative = "two.sided")
```


